Transient Corotating Clumps Around Adolescent Low-Mass Stars From Four Years of TESS

### 2023-09-07 paper/ms.tex
update ms
Co-authored-by: Lynne Hillenbrand <lah@astro.caltech.edu>

% 250 word max!
\begin{abstract}
  %
  Complex periodic variables (CPVs) are stars that exhibit sharp and
  periodic optical light curves.  Previous work has indicated that such stars
  tend to be disk-free pre-main-sequence M dwarfs with rotation
  periods of 0.2--2 days.  To advance understanding of these enigmatic
  objects, we performed a blind search using TESS 2-minute data of
  \nstarssearched\ K and M dwarfs with $T$$<$16 and $d$$<$150\,pc.  We
  found \ngoods\ high-quality CPVs.  The new objects include the
  brightest ($T$$\approx$9.5), closest ($d$$\approx$20\,pc), and
  oldest ($\approx$200\,Myr) CPVs known.  All objects but one turned
  out to be likely members of stellar associations, which enabled
  age-dating.  An exceptional new object, LP 12-502, showed up to
  eight flux dips per cycle; some of the LP 12-502 dips coexisted with slightly
  different periods, and the shortest-duration LP 12-502 dips matched the expected
  timescale for transiting small bodies at the corotation radius.
  Broadly, our search confirms
  that CPVs are mostly young ($\lesssim$150\,Myr) and low-mass
  ($\lesssim$0.4\,$M_\odot$).  The flux dips that embody the class
  have lifetimes of $\approx$100 cycles, and the dips collapse in
  response to stellar flares once every few months.  Corotating
  concentrations of gas or dust remains the only plausible explanation
  known.  The gas or dust is probably entrained by the star's magnetic
  field, and the sharp features could result from a multipolar field
  topology, a hypothesis supported by correspondences between the
  light curves of CPVs and of B stars known to have multipolar
  magnetic fields.
  %
\end{abstract}

@@

Table~\ref{tab:thetable} summarizes each of the sources of
binarity information into a single bitwise column.  
We present the detailed results concerning binarity in
Section~\ref{subsec:resultsbinarity}, and a summary discussion in
Section~\ref{subsec:discbinary}.

@@

\begin{figure*}[!t]
	\begin{center}
		\centering
		\includegraphics[width=0.98\textwidth]{f8.pdf}
		\vspace{-0.3cm}
		\caption{
			{\bf Evolution of LP 12-502} ($P$=18.5611\,h) at fixed period and
      epoch over three years.  Each panel shows one (averaged) TESS
			orbit; small text denotes relative cycle number.  There are 200
			binned black points per cycle.  The TESS pointing law dictates
			the large time gaps between cycles 64-248, 315-1233, and
			1264-1410; larger gaps tend to yield larger shape changes.  The
			dips usually evolve over tens to hundreds of cycles.  However
			cycles 1233-1264 show a dip that switched from a depth and
			duration of 3\% and 3\,hr to 0.3\% and 1\,hr over less than one
			cycle (cf.~Figure~\ref{fig:lplc}).
		}
		\label{fig:lp}
	\end{center}
\end{figure*}

@@

The closest newly-recognized CPV in our catalog is DG~CVn (TIC~368129164),
a member of AB Dor at $d$$=$18\,pc.  
The three brightest CPVs are DG~CVn
($T$=9.3), TIC~405754448 ($T$=9.6), and TIC~167664935 ($T$=10.3).  The
shortest period, 3.64\,hr, belongs to TIC~201789285.  The longest
period, 37.9\,hr, belongs to TIC~405910546.  Based on the Gaia$_{\rm DR3}$ RV
scatter, the latter source may
turn out to be an eclipsing binary; if so, the longest-period CPV in our
catalog would be TIC~193831684 (31.0\,hr).  By definition, we required
the periods to be below 48\,hr.

@@

For CPVs, binarity seems to mostly provide a mix of
observational nuisances and curiosities.  The
nuisances include
astrophysical false positives with two beating rapidly
rotating stars, as well as uncertainty about which
star is producing the CPV signal in binary systems.  Our two candidate K dwarf CPVs 
suffer from the latter concern (see Section~\ref{subsec:resultsbinarity}).

Curiosities include the four binary systems that are now known to each host
two separate CQVs.
CQVs are sufficiently rare that such systems may have 
physical implications.
For instance, recent work has shown that the orbits of binaries closer than
$\lesssim$700 AU tend to be aligned with their planetary systems
\citep[e.g.][]{2022AJ....163..207C}.  If we assume that observing CPV
variability requires high inclinations relative to the line of sight,
and that the inclinations in binaries are correlated, then we would
expect the detection of one CPV in a binary system to raise the
probability that the other star is a CPV.  The limitations of the
current catalog prevent further exploration of this issue, but it
might be interesting for future study.

@@

While CPV periods appear to remain constant to within measurement
precision over thousands of cycles, the light curve shapes evolve over
typical timescales of 10 to 1{,}000 cycles (e.g.
Figures~\ref{fig:evoln} and~\ref{fig:lpriver0}).  The CPVs are
therefore quasiperiodic, with quality factors of $Q$$\approx$100.
While this observation is consistent with the analyses by
\citet{2022AJ....163..144G} and \citet{2023ApJ...945..114P}, the fact
that all CPV dips seem to be transient over timescales of more than
1{,}000 cycles (Figure~\ref{fig:evoln}) marks a qualitative departure
from the ``persistent'' {\it vs.} ``transient'' flux dip distinction
previously described by \citet{2017AJ....153..152S}.

@@

\subsection{Dip asymmetries and dust geometries}

@@

\subsection{Dust vs. gas}

Both the dust clump and the gas prominence scenario invoke clumps
of material at the corotation radius;
the main property that we believe distinguishes the two ideas is 
the composition of the material.

@@

An instructive point of comparison is the rapidly rotating magnetic B
star, $\sigma$~Ori~E, which shows dips that are deeper in the blue
than in the red \citep{1977ApJ...216L..31H}.  Photometric and
spectroscopic observations of this star have been understood in terms
of a warped torus of corotating circumstellar material
\citep{1978ApJ...224L...5L,1985Ap&SS.116..285N,2005ApJ...630L..81T}.
The circumstellar material is unlikely to be dust, which would
sublimate quickly at the distance of the torus from the
star.\footnote{\citet{2019ApJ...876..127Z} explored the sublimation
timescales for a canonical CPV with $M_\star=0.2\,M_\odot$,
$R_\star=0.3\,R_\odot$, and $T_{\rm eff}=3200\,{\rm K}$.  They found
that non-shielded, generic silicate dust mixture \citep{1985ApJS...57..587D} with a
single size of 0.1\,$\mu$m reached the $\approx$1500\,K sublimination
temperature at $\approx$3\,$R_\star$.  This suggests that dust
sublimation could be an important effect even for CPVs. }  The opacity
source for $\sigma$~Ori~E and its analogs is instead thought to be
bound-free absorption by neutral hydrogen \citep{1985Ap&SS.116..285N},
although to our best knowledge direct evidence for this conclusion has
yet to be acquired.  Separate and smaller-amplitude continuum flux
brightenings in $\sigma$~Ori~E may also come from electrons scattering
photospheric light toward the observer when the clouds are not
transiting \citep{2022MNRAS.511.4815B}.

@@

In the context of disk-driven planet migration, the stopping location
for the innermost planet is set by the location of the protoplanetary
disk's truncation radius \citep[e.g.][and references therein]{2018haex.bookE.142I}.  The 
truncation radius is traditionally calculated by equating the magnetic
pressure from the stellar magnetosphere with the ram pressure of the
inflowing gas.  As it happens, the truncation radius is close to the
corotation radius for low accretion rates
\citep[e.g.][]{2016ARA&A..54..135H,2022MNRAS.510.5246L}.  These considerations
invite us to imagine one or more planets migrating inward due to gas
drag, and arriving at $\approx$5-10 stellar
radii before the disk is depleted.

@@

		\vspace{-0.35cm}
		\subfloat{
			\includegraphics[width=0.48\textwidth]{f10b.pdf}
		}

		\vspace{-0.35cm}
		\subfloat{
			\includegraphics[width=0.48\textwidth]{f10c.pdf}
		}

@@

      collapsing instantaneously, and are almost always linked with

>>>>

% 250 word max!
\begin{abstract}
  %
  Complex periodic variables (CPVs) are stars that exhibit sharp and
  periodic optical light curves.  Previous work has indicated that such stars
  tend to be disk-free pre-main-sequence M dwarfs with rotation
  periods of 0.2--2 days.  To advance understanding of these enigmatic
  objects, we performed a blind search using TESS 2-minute data of
  \nstarssearched\ K and M dwarfs with $T$$<$16 and $d$$<$150\,pc.  We
  found \ngoods\ high-quality CPVs.  The new objects include the
  brightest ($T$$\approx$9.5), closest ($d$$\approx$20\,pc), and
  oldest ($\approx$200\,Myr) CPVs known.  All objects but one turned
  out to be likely members of stellar associations, which enabled
  age-dating.  An exceptional new object, LP 12-502, showed up to
  eight flux dips per cycle; some of the LP 12-502 dips coexisted with slightly
  different periods, and the shortest-duration LP 12-502 dips matched the expected
  timescale for transiting small bodies at the corotation radius.
  Broadly, our search confirms
  that CPVs are mostly young ($\lesssim$150\,Myr) and low-mass
  ($\lesssim$0.4\,$M_\odot$).  The flux dips that embody the class
  have lifetimes of $\approx$100 cycles, and the dips collapse following
  stellar flares once every few months.  Corotating
  concentrations of gas or dust remains the only plausible explanation
  known.  The gas or dust is probably entrained by the star's magnetic
  field, and the sharp features could result from a multipolar field
  topology, a hypothesis supported by correspondences between the
  light curves of CPVs and of B stars known to have multipolar
  magnetic fields.
  %
\end{abstract}

@@

Table~\ref{tab:thetable} summarizes each of the sources of
binarity information into a single bitwise column.  
We present detailed results concerning binarity in
Section~\ref{subsec:resultsbinarity}, and a summary discussion in
Section~\ref{subsec:discbinary}.

@@

\begin{figure*}[!t]
	\begin{center}
		\centering
		\includegraphics[width=0.99\textwidth]{f8.pdf}
		\vspace{-0.3cm}
		\caption{
			{\bf Evolution of LP 12-502} ($P$=18.5611\,h) at fixed period and
      epoch over three years.  Each panel shows one (averaged) TESS
			orbit; small text denotes relative cycle number.  There are 200
			binned black points per cycle.  The TESS pointing law dictates
			the large time gaps between cycles 64-248, 315-1233, and
			1264-1410; larger gaps tend to yield larger shape changes.  The
			dips usually evolve over tens to hundreds of cycles.  However
			cycles 1233-1264 show a dip that switched from a depth and
			duration of 3\% and 3\,hr to 0.3\% and 1\,hr over less than one
			cycle (cf.~Figure~\ref{fig:lplc}).
		}
		\label{fig:lp}
	\end{center}
\end{figure*}

@@

The closest CPV in our catalog is DG~CVn (TIC~368129164),
a member of AB Dor at $d$$=$18\,pc.  
The three brightest CPVs are DG~CVn
($T$=9.3), TIC~405754448 ($T$=9.6), and TIC~167664935 ($T$=10.3).  The
shortest period, 3.64\,hr, belongs to TIC~201789285.  The longest
period, 37.9\,hr, belongs to TIC~405910546.  Based on the Gaia$_{\rm DR3}$ RV
scatter, the latter source may
turn out to be an eclipsing binary; if so, the longest-period CPV in our
catalog would be TIC~193831684 (31.0\,hr).  By definition, we required
the periods to be below 48\,hr.

@@

For CPVs, binarity seems to mostly provide a mix of
nuisances and curiosities.  The
nuisances include
astrophysical false positives with two beating rapidly
rotating stars, as well as uncertainty about which
star produces the CPV signal in binary systems.  Our two candidate K dwarf CPVs 
suffer from this latter concern (see Section~\ref{subsec:resultsbinarity}).

Curiosities include the four binary systems that are now known to each host
two separate CQVs.
CQVs are sufficiently rare that such systems may have 
physical implications.
Recent work has shown that the orbits of binaries closer than
$\lesssim$700 AU tend to be aligned with their planetary systems
\citep[e.g.][]{2022AJ....163..207C}.  If we assume that observing CPV
variability requires high line-of-sight inclinations,
and that the inclinations in binaries are correlated, then we would
expect the detection of one CPV in a binary system to raise the
probability that the other star is a CPV.  The limitations of the
current catalog prevent further exploration of this issue, but it
might be interesting for future study.

@@

While CPV periods appear to remain fixed over thousands of cycles, the light curve shapes evolve over
typical timescales of 10 to 1{,}000 cycles (e.g.
Figures~\ref{fig:evoln} and~\ref{fig:lpriver0}).  Although we refer
to them as ``periodic'', the CPVs are
therefore actually quasiperiodic, with quality factors of $Q$$\approx$100.
While this observation is consistent with the analyses by
\citet{2022AJ....163..144G} and \citet{2023ApJ...945..114P}, the fact
that all CPV dips seem to be transient over timescales of more than
1{,}000 cycles (Figure~\ref{fig:evoln}) marks a qualitative departure
from the ``persistent'' {\it vs.} ``transient'' flux dip distinction
previously described by \citet{2017AJ....153..152S}.

@@

\subsection{Dip asymmetries?}

@@

\subsection{What causes the CPV phenomenon: dust vs.\ gas}

Both the dust clump and the gas prominence scenarios (Figure~\ref{fig:f1} and Section~\ref{sec:intro})
invoke clumps
of material at the corotation radius;
the main property that we believe distinguishes the two ideas is 
the composition of the material.

@@

An instructive point of comparison is the rapidly rotating magnetic B
star, $\sigma$~Ori~E, which shows dips that are deeper in the blue
than in the red \citep{1977ApJ...216L..31H}.  Photometric and
spectroscopic observations of this star have been understood in terms
of a warped torus of corotating circumstellar material
\citep{1978ApJ...224L...5L,1985Ap&SS.116..285N,2005ApJ...630L..81T}.
The circumstellar material is unlikely to be dust, which would
sublimate quickly at the distance of the torus from the
star.\footnote{\citet{2019ApJ...876..127Z} explored the sublimation
timescales for a canonical CPV with $M_\star=0.2\,M_\odot$,
$R_\star=0.3\,R_\odot$, and $T_{\rm eff}=3200\,{\rm K}$.  They found
that non-shielded, generic silicate dust mixture \citep{1985ApJS...57..587D} with a
single size of 0.1\,$\mu$m reached the $\approx$1500\,K sublimation
temperature at $\approx$3\,$R_\star$.  This suggests that dust
sublimation could be an important effect even for CPVs. }  The opacity
source for $\sigma$~Ori~E and its analogs is instead thought to be
bound-free absorption by neutral hydrogen \citep{1985Ap&SS.116..285N},
although to our best knowledge direct evidence for this conclusion has
yet to be acquired.  Separate and smaller-amplitude continuum flux
brightenings in $\sigma$~Ori~E may also come from electrons scattering
photospheric light toward the observer when the clouds are not
transiting \citep{2022MNRAS.511.4815B}.

@@

In the context of disk-driven planet migration, the stopping location
for the innermost planet is set by the location of the protoplanetary
disk's truncation radius \citep[e.g.][and references therein]{2018haex.bookE.142I}.  The 
truncation radius is traditionally calculated by equating the magnetic
pressure from the stellar magnetosphere with the ram pressure of the
inflowing gas.  As it happens, the truncation radius is close to the
corotation radius for low accretion rates
\citep[e.g.][]{2015SSRv..191..339R,2022MNRAS.510.5246L}.  These considerations
invite us to imagine one or more planets migrating inward due to gas
drag, and arriving at $\approx$5-10 stellar
radii before the disk is depleted.

@@

		\vspace{-0.34cm}
		\subfloat{
			\includegraphics[width=0.48\textwidth]{f10b.pdf}
		}

		\vspace{-0.34cm}
		\subfloat{
			\includegraphics[width=0.48\textwidth]{f10c.pdf}
		}

@@

      collapsing instantaneously, and
      occur once every few months for
      both LP~12-502 and TIC~300651846.
      They are almost always linked with
